---
title: "Modeling with XGBoost"
author: "Elias Grant 306445678"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Libraries and Data
```{r}
library(tidyverse)
library(xgboost)
library(caret)
data <- read_rds("stuff_data.rds")
```

# Feature Engineering

### Adjust release location, approach angle, and horizontal break to account for pitcher handedness
```{r}
# Horizontal release location
data <- data |>
  mutate(RelSide2 = ifelse(PitcherThrows == "Left", -1 * RelSide, RelSide))

# Horizontal break
data <- data |>
  mutate(HorzBreak2 = ifelse(PitcherThrows == "Left", -1 * HorzBreak, HorzBreak))

# Spin induced horizontal break
data <- data |>
  mutate(pfxx2 = ifelse(PitcherThrows == "Left", -1 * pfxx, pfxx))

# Horizontal approach angle
data <- data |>
  mutate(HorzApprAngle2 = ifelse(PitcherThrows == "Left", -1 * HorzApprAngle, HorzApprAngle))
```

### Create adjusted velocity and movement metrics relative to fastball
```{r}
# Velocity Differential
fb_velo <- data |>
  group_by(Pitcher) |>
  filter(TaggedPitchType == "Fastball" | TaggedPitchType == "FourSeamFastBall") |>
  summarize(avg_fbvelo = mean(RelSpeed, na.rm = TRUE))

data <- data |>
  left_join(fb_velo, by = "Pitcher")

data <- data |>
  mutate(speed_diff = RelSpeed - avg_fbvelo)

# Movement Differential
fb_break <- data |>
  group_by(Pitcher) |>
  filter(TaggedPitchType == "Fastball" | TaggedPitchType == "FourSeamFastBall") |>
  summarize(avg_fbbreak = mean(pfxx2, na.rm = TRUE))

data <- data |>
  left_join(fb_break, by = "Pitcher")

data <- data |>
  mutate(break_diff = pfxx2 - avg_fbbreak)

# Remove excess tibbles
rm(fb_break, fb_velo)


# Export dataset 
saveRDS(object = data, file = "stuffplus_data.rds")
```


# Modeling XGBoost
```{r}
# Data Preparation
data <- data |>
  select(c("Pitcher", "TaggedPitchType", "RelSpeed", "SpinRate", "RelHeight", 
           "RelSide2", "Extension", "InducedVertBreak", "pfxx2", "VertApprAngle", 
           "HorzApprAngle2", "speed_diff", "break_diff", "linear_weight", "lw_event")) |>
  drop_na()

# Convert lw_event to a 0-indexed numeric factor required for XGBoost multi-class
data$lw_event <- as.factor(data$lw_event)
data$class_label <- as.numeric(data$lw_event) - 1
num_classes <- length(unique(data$class_label))

set.seed(10293939)
train_index <- createDataPartition(data$linear_weight, p = 0.8, list = FALSE)

train_df <- data[train_index,]
test_df <- data[-train_index,]

feature_cols <- c("RelSpeed", "SpinRate", "RelHeight", "RelSide2", "Extension", 
                  "InducedVertBreak", "pfxx2", "VertApprAngle", "HorzApprAngle2", 
                  "speed_diff", "break_diff")

features1 <- as.matrix(train_df[, feature_cols])
features2 <- as.matrix(test_df[, feature_cols])


# REGRESSION MODEL
label1_reg <- train_df$linear_weight
label2_reg <- test_df$linear_weight

train_reg <- xgb.DMatrix(data = features1, label = label1_reg)
test_reg  <- xgb.DMatrix(data = features2, label = label2_reg)

tune_grid <- expand.grid(
  eta = c(0.01, 0.05, 0.1),
  max_depth = c(3, 5, 7),
  subsample = c(0.8, 1.0),
  colsample_bytree = c(0.8, 1.0)
)

best_rmse <- Inf
best_params_reg <- list()

cat("Starting Grid Search for Regression Model...\n")

for (i in 1:nrow(tune_grid)) {
  params <- list(
    objective = "reg:squarederror",
    eval_metric = "rmse",
    eta = tune_grid$eta[i],
    max_depth = tune_grid$max_depth[i],
    subsample = tune_grid$subsample[i],
    colsample_bytree = tune_grid$colsample_bytree[i]
  )
  
  set.seed(10293939)
  
  mdl <- xgb.train(
    params = params,
    data = train_reg,
    nrounds = 200, 
    evals = list(val = test_reg), 
    early_stopping_rounds = 10,
    verbose = 0
  )
  
  # Manually calculate RMSE
  preds <- predict(mdl, test_reg)
  current_rmse <- sqrt(mean((preds - label2_reg)^2))
  
  if (current_rmse < best_rmse) {
    best_rmse <- current_rmse
    best_params_reg <- params
  }
}

cat("Best Regression RMSE:", best_rmse, "\n")
print(best_params_reg)

# Train Final Regression Model
model_reg_final <- xgb.train(
  params = best_params_reg,
  data = train_reg,
  nrounds = 500,
  evals = list(train = train_reg, test = test_reg), 
  early_stopping_rounds = 10,
  print_every_n = 50
)

# CLASSIFICATION MODEL

# Create DMatrix objects for classification
label1_class <- train_df$class_label
label2_class <- test_df$class_label

train_class <- xgb.DMatrix(data = features1, label = label1_class)
test_class  <- xgb.DMatrix(data = features2, label = label2_class)

# Reformat linear weights to a vector, 
# Ensure run-value is constant for each classification
class_run_values <- train_df |>
  group_by(class_label) |>
  summarize(avg_rv = mean(linear_weight, na.rm = TRUE)) |>
  arrange(class_label) |>
  pull(avg_rv)

params_class <- list(
  objective = "multi:softprob",
  eval_metric = "mlogloss",
  num_class = num_classes,
  eta = 0.05,
  max_depth = 5
)

# Train Classification Model
model_class <- xgb.train(
  params = params_class,
  data = train_class,
  nrounds = 1000,
  evals = list(train = train_class, test = test_class), 
  early_stopping_rounds = 10,
  print_every_n = 50
)

# Predictions and Evaluation

# Predict probabilities on test set
pred_probs <- predict(model_class, test_class)

# Calculate Expected Run Value (xRV) via matrix multiplication
test_df$predicted_xRV <- as.vector(pred_probs %*% class_run_values)

# Compare regression predictions vs classification xRV predictions
test_df$predicted_reg <- predict(model_reg_final, test_reg)

# View results
head(test_df[, c("linear_weight", "predicted_reg", "predicted_xRV")])
```

### Feature Importance 
```{r}
# Generate the importance matrices mapping your feature columns to the models
importance_reg <- xgb.importance(feature_names = feature_cols, model = model_reg_final)
importance_class <- xgb.importance(feature_names = feature_cols, model = model_class)

# Print the dataframes to the console
cat("\nTop Features: Regression Model\n")
print(head(importance_reg, 5))

cat("\nTop Features: Classification Model\n")
print(head(importance_class, 5))

# Plot the feature importance for visual comparison
xgb.plot.importance(
  importance_matrix = importance_reg, 
  top_n = 10, 
  measure = "Gain",
  main = "Regression Predictor Importance"
)

xgb.plot.importance(
  importance_matrix = importance_class, 
  top_n = 10, 
  measure = "Gain",
  main = "Classification Predictor Importance"
)
```

### Stuff+ Normalization
```{r}
# Calculate Stuff+ and group by pitcher

# Calculate league averages and standard deviations from the test set
mean_xRV <- mean(test_df$predicted_xRV, na.rm = TRUE)
sd_xRV <- sd(test_df$predicted_xRV, na.rm = TRUE)

mean_reg <- mean(test_df$predicted_reg, na.rm = TRUE)
sd_reg <- sd(test_df$predicted_reg, na.rm = TRUE)

# Score every individual pitch on the 100 scale
# Formula: 100 - (Z-score * 10). 
# Note: subtract the z-score because a lower run value denotes better pitcher performance.
test_df <- test_df |>
  mutate(
    pitch_StuffPlus_xRV = 100 - ((predicted_xRV - mean_xRV) / sd_xRV) * 10,
    pitch_StuffPlus_Reg = 100 - ((predicted_reg - mean_reg) / sd_reg) * 10
  )

# Group by Pitcher and summarize
pitcher_leaderboard <- test_df |>
  group_by(Pitcher) |>
  summarize(
    Pitches = n(),
    # Calculate the average Stuff+ score for the pitcher
    StuffPlus_xRV = round(mean(pitch_StuffPlus_xRV, na.rm = TRUE), 1),
    StuffPlus_Reg = round(mean(pitch_StuffPlus_Reg, na.rm = TRUE), 1)
  ) |>
  # Filter out low-sample pitchers
  filter(Pitches >= 20) |> 
  # Sort by the Classification xRV model's Stuff+ descending
  arrange(desc(StuffPlus_xRV))

# View the top 10 pitchers by Stuff+
cat("\n--- Top 10 Pitchers by Stuff+ (xRV Model) ---\n")
print(head(pitcher_leaderboard, 10))

# View the bottom 10 pitchers by Stuff+
cat("\n--- Bottom 10 Pitchers by Stuff+ (xRV Model) ---\n")
print(tail(pitcher_leaderboard, 10))
```

Overall, after cross-checking with Baseball Reference for the actual results of the top and bottom 10 pitchers, initial results are promising. Higher sample size is a must-have, but the work so far is a solid foundation to build on
